Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-05-13 unimodular-covers.tex
wrote draft of cayley section

So, the questions to look at are:

1) Does every lattice $3$-parallelepiped have the circumscribed parallelepiped property? I think yes.

2) Does the Cayley embedding of a $P$ and a $Q$ such that one normal fan refines the other have the circumscribed parallelepiped property? Probably not in general, but perhaps yes in some cases; probably yes if one of them is a Minkowski summand of the other.

2’) Even if the answer to (2) is negative, the following can be enough to prove unimodular cover: find a triangulation of $Cayley(P,Q)$ in which all the tetrahedra with an edge in $P$ and an edge in $Q$ have the

3) The long shot: does every smooth 3-polytope have the circumscribed parallelepiped property?

Paco

\section{nef Cayley polytopes}
\label{sec:cayley}

>>>>

\begin{question}
Does every smooth 3-polytope have the circumscribed parallelepiped property? By the above remark, if the answer is yes then every smooth $3$-polytope has a unimodular cover.
\end{question}

\section{nef Cayley sums}
\label{sec:cayley}

Throughout this section let $P$ and $Q$ be two lattice polytopes in $\R^d$. We do not require them to be full-dimensional, but we assume their Minkowski sum is. Remember that the Minkowski sum $P+Q$ of $P$ and $Q$ is defined as:
\[
P + Q := \{ p+q \in \R^d: p\in P, q \in Q\}.
\]
The \emph{Cayley sum} of $P$ and $Q$ is defined as
\[
\cayley (P,Q) = \conv( P\times\{0\} \cup Q\times \{1\}) \in \R^{d+1}.
\]
The assumption that $P+Q$ is $d$-dimensional is equivalent to $\cayley (P,Q)$ is a $(d+1)$-dimensional.

We introduce the following two concepts.

\begin{definition}
In the above conditions, we say that $Q$ \emph{is nef for $P$} if the normal fan of $Q$ refines that of $P$. If this happens, we say that $\cayley(P,Q)$ is a \emph{nef Cayley sum}.

We say that the pair $(P,Q)$ is \emph{mixed IDP} if 
\paco{``mixed IDP'' is not the right property, since this is weaker than IDP}
\[
(P\cap \Z^d) + (Q\cap \Z^d) = (P+Q)\cap \Z^d.
\]
\end{definition}

In this language, Problems 3 and 4 i \cite{Oda1997} translate to

\begin{question}
If $Q$ is nef for $P$, is the pair $(P,Q)$ mixed IDP?
\end{question}

In dimension two, the answer is always yes by \cite{HNPS2008}. In dimension three it is easy to construct counter-examples (e.g., let $P=Q$ be the empty simplex of volume two), but the case where $P$ is smooth is open, and attributed as a conjecture of Oda.

\begin{proposition}
If $\cayley(P,Q)$ is IDP then the pair $(P,Q)$ is mixed IDP.
\end{proposition}

The so-called \emph{Cayley Trick} is the isomorphism
\[
2\cayley(P, Q) \cap (\R^d\times \{1\}) \cong P+Q,
\]
which induces the following canonical bijections:
\begin{align*}
\text{polyhedral subdivisions of $\cayley(P,Q)$} &\leftrightarrow \text{mixed subdivisions of $P + Q$}\\
\text{triangulations of $\cayley(P,Q)$} &\leftrightarrow \text{fine mixed subdivisions of $P + Q$}\\
\text{unimodular simplices in $\cayley(P,Q)$} &\leftrightarrow \text{unimodular prod-simplices in $P + Q$}.
\end{align*}
See \cite{DLRS2010} for more details on the Cayley Trick and on triangulations and polyhedral subdivisions of polytopes.
In fact these bijections can be taken as definitions of the objects in the right-hand sides. In particular, 
we call \emph{prod-simplices} in $P+Q$ the Minkowski sums $T_1+T_2$ where $T_1\subset P$ and $T_2\subset Q$ are simplices with complementary affine spans. A prod-simplex is \emph{unimodular} if the edge vectors from a vertex of $T_1$ and from a vertex of $T_2$ form a unimodular basis.

%As a consequence, we also have that:
%\begin{align*}
%\text{$\cayley(P,Q)$ has a unimodular cover} &\leftrightarrow \text{$P+Q$ can be covered by unimodular prod-simplices}.
%\end{align*}

Our proof of \Cref{thm:cayley} is based on the following result:

\begin{lemma}
\label{lemma:cayley}
Let $P$ and $Q$ be two lattice polygons with $Q$ nef with respect to $P$. Let $p_1p_2 \subset P$ and $q_1q_2\subset Q$ be two primitive and non-parallel segments, and assume that the prod-simplex (parallelogram)  $p_1p_2 + q_1q_2$ is not unimodular. Then,
at least one of the strips
\paco{add picture for these strips, and perhaps for the proof}
\[
(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle) \cap P
\qquad \text{and} \qquad
(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle) \cap Q
\]
has a lattice point in its interior.
\end{lemma}

\begin{proof}
To DO
\end{proof}

\begin{proof}[Proof of \Cref{thm:cayley}]
TO DO 
\end{proof}
